A spiral galaxy's mass distribution uncovered through lensing and dynamics

### 2015-08-20 docs/model_JAM_v1.01.tex
Started writing section on JAM modelling with surface brightness MGE. Updated plot on model G enclosed mass.
Wrote section on rotation curve recovery.

\subsection{Jeans Axisymmetric Modelling (JAM)}

Jeans axisymmetric models (JAM) assume galaxies (a) to be collisionless, i.e. the collisionless Boltzmann equation for the distribution function $f(\vect{x},\vect{v},t)$ has to be satisfied ($\frac{\diff f(\vect{x},\vect{v},t}{\diff t} = 0$), (b) in a steady state ($\frac{\partial}{\partial t} = 0$), (c) axisymmetric (best described in cylindrical coordinates $(R,z,\phi)$ and $\frac{\partial}{\partial \phi} = 0$). From this follow the axisymmetric Jeans equations as the vector-valued first moment of the Boltzmann equation, i.e.
\begin{equation*}
\int \frac{\diff f}{\diff t} \Diff3 v = 0.
\end{equation*}
To be able to solve the Jeans equations, additional assumptions about the velocity ellipsoid tensor $\langle v_i v_j\rangle$ have to be made. We follow \citet{Cap08} and assume firstly, that the galaxy's velocity ellipsoid is aligned with the cylindrical coordinate system, i.e. $\langle v_i v_j\rangle = 0$ for $i\neq j$. Secondly, we assume a constant ratio between the radial and vertical 2nd velocity moments, $\beta_z \equiv 1 - \langle v_z^2 \rangle / \langle v_R^2\rangle$. This reduces the Jeans equations to two equations for $\langle v_z^2 \rangle$ and $\langle v_\phi^2 \rangle$, that can be solved by means of one integration,
\begin{eqnarray*}
n \langle v_z^2 \rangle (R,z) &=& \int_0^\infty n \frac{\partial \Phi}{\partial z} \diff z\\
n \langle v_\phi^2 \rangle (R,z) &=& R \frac{\partial}{\partial R} \left( \frac{n \langle v_z^2 \rangle}{1-\beta_z} \right) + \frac{n \langle v_z^2 \rangle}{1 - \beta_z} + R n \frac{\partial \Phi}{\partial R},
\end{eqnarray*}
where $n(\vect{x}) = \int f(\vect{x},\vect{v}) \diff3 v$ is the number density of tracers and $\Phi(\vect{x})$ the galaxy's gravitational potential, generated by the mass density $\rho(\vect{x})$ via Poisson's equation.
\\The JAM modelling approach by \citet{Cap08} makes use of expressing the tracer density and the mass density as MGEs (see also [TO DO: REF: Emsellem et al. 19994]). The density of stellar tracers is assumed to be proportional to the observed and deprojected brightness distribution $\nu(R,z)$ in Eq. (\ref{eq:deprojMGE}). The mass density consists of several sets of MGEs: One MGE, that is usually taken to be $\nu(R,z)$ multiplied by a constant stellar mass-to-light ratio $\Upsilon_*$, describes the distribution of stellar mass in the galaxy. To mimick gradients of stellar mass-to-light ratio, each Gaussian could be assigned its own $\Upsilon_{*,i}$. To add a Navarro-Frenck-White (NFW) [TO DO: REF] dark matter halo component, a MGE generated from a fit to a NFW profile can be added to the stellar component. [TO DO: continue on p. 74]

\paragraph{Data comparison.} As data we use stellar line-of-sight rotation velocities $v_\text{rot} \equiv \langle v_\text{los} \rangle$ [TO DO: consistent, los or rot] and velocity dispersions $\sigma$ as described in \S\ref{sec:data}. The JAM models give us a prediction for the second line-of-sight velocity moment $v_\text{los}$. The root mean square (rms) line-of-sight velocity $v_\text{rms}$ allows a data-model comparison by relating theses velocities according to 
\begin{equation*}
v_\text{rms}^2 = \langle v_\text{los}^2 \rangle = v_\text{rot}^2 + \sigma^2.
\end{equation*}
The model in Eq. [TO DO] predicts the intrinsic $\langle v_\text{los}^2 \rangle$ at a given position on the sky, which have then to be modified to model the mode of observation, to be comparable to the measurements. The measured $v_\text{rms}$ is a light-weighted mean for a pixel along the long-slit of the spectrograph, with height $L_y = 1$ arcsec [TO DO: REF: Dutton 2013] and a certain given extent in along the major axis, $L_x$, i.e. for a rectangular aperture
\begin{equation*}
\text{AP}(x,y) = \left\{ \begin{array}{ll} 1 & \text{for } -\frac{L_x}{2} \leq x \leq + \frac{L_x}{2} \text{ and } - \frac{L_y}{2} \leq y \leq + \frac{L_y}{2}  \\ 0 & \text{otherwise} \end{array} \right. .
\end{equation*}
The light arriving at the spectrograph itself was subject to seeing, i.e. a Gaussian 
\begin{equation*}
\text{PSF}(x,y)=\mathscr{N}(0,FWHM/2\sqrt{2\ln2})
\end{equation*}
with FWHM=1.1 arcsec  [TO DO: REF: Dutton 2013]. The model predictions have therefore to be convolved with the convolution kernel
\begin{eqnarray*}
K(x,y) &=& (\text{PSF} \ast \text{AP})(x,y) \\
&=& \frac{1}{4} \prod_{u \in \{x,y\}} \left[ \text{erf}\left( \frac{L_u/2 - u}{\sqrt{2}\sigma_\text{seeing}}\right) + \text{erf} \left( \frac{L_u/2 + u}{\sqrt{2} \sigma_\text{seeing}} \right) \right]
\end{eqnarray*}
and weighted by the surface brightness $I(x,y)$ [TO DO: primed x and y or not????]
\begin{eqnarray*}
I_\text{obs} &=& I \ast K\\
\langle v_\text{los}^2 \rangle |_\text{obs} &=& \frac{(I \langle v_\text{los}^2\rangle) \ast K}{I_\text{obs}}.
\end{eqnarray*}
If provided with with the convolution kernel, the JAM code by [TO DO: REF: Cappellari 2008] [TO DO: reference code] performs the convolution numerically. We set $L_x = 0.21$ arcsec as the width of the model pixel, and get a prediction for the actual measurements in bins of width 0.63, 1.26 and 1.89 arcsec [TO DO: REF: Dutton 2013] as light-weighted mean from each 3, 6 and 9 model pixels.

>>>>

\subsection{Jeans Axisymmetric Modelling (JAM)} \label{sec:model_JAM}

Jeans axisymmetric models (JAM) assume galaxies (a) to be collisionless, i.e. the collisionless Boltzmann equation for the distribution function $f(\vect{x},\vect{v},t)$ has to be satisfied ($\frac{\diff f(\vect{x},\vect{v},t}{\diff t} = 0$), (b) in a steady state ($\frac{\partial}{\partial t} = 0$), (c) axisymmetric (best described in cylindrical coordinates $(R,z,\phi)$ and $\frac{\partial}{\partial \phi} = 0$). From this follow the axisymmetric Jeans equations as the vector-valued first moment of the Boltzmann equation, i.e.
\begin{equation*}
\int \frac{\diff f}{\diff t} \Diff3 v = 0.
\end{equation*}
To be able to solve the Jeans equations, additional assumptions about the velocity ellipsoid tensor $\langle v_i v_j\rangle$ have to be made. We follow \citet{Cap08} and assume firstly, that the galaxy's velocity ellipsoid is aligned with the cylindrical coordinate system, i.e. $\langle v_i v_j\rangle = 0$ for $i\neq j$. Secondly, we assume a constant ratio between the radial and vertical 2nd velocity moments, $\beta_z \equiv 1 - \langle v_z^2 \rangle / \langle v_R^2\rangle$. This reduces the Jeans equations to two equations for $\langle v_z^2 \rangle$ and $\langle v_\phi^2 \rangle$, that can be solved by means of one integration,
\begin{eqnarray*}
n \langle v_z^2 \rangle (R,z) &=& \int_0^\infty n \frac{\partial \Phi}{\partial z} \diff z\\
n \langle v_\phi^2 \rangle (R,z) &=& R \frac{\partial}{\partial R} \left( \frac{n \langle v_z^2 \rangle}{1-\beta_z} \right) + \frac{n \langle v_z^2 \rangle}{1 - \beta_z} + R n \frac{\partial \Phi}{\partial R},
\end{eqnarray*}
where $n(\vect{x}) = \int f(\vect{x},\vect{v}) \diff3 v$ is the number density of tracers and $\Phi(\vect{x})$ the galaxy's gravitational potential, generated by the mass density $\rho(\vect{x})$ via Poisson's equation.
\\The JAM modelling approach by \citet{Cap08} makes use of expressing the tracer density and the mass density as MGEs (see also \citet{1994A&A...285..723E}). The density of stellar tracers is assumed to be proportional to the observed and deprojected brightness distribution $\nu(R,z)$ in Eq. (\ref{eq:deprojMGE}). The mass density consists of several sets of MGEs: One MGE, that is usually taken to be $\nu(R,z)$ multiplied by a constant stellar mass-to-light ratio $\Upsilon_*$, describes the distribution of stellar mass in the galaxy. To mimick gradients of stellar mass-to-light ratio, each Gaussian could be assigned its own $\Upsilon_{*,i}$. To add a Navarro-Frenck-White (NFW) \citep{Navarro+1995c,NFW96} dark matter halo component, a MGE generated from a fit to a NFW profile can be added to the stellar component. [TO DO: continue on p. 74]

\paragraph{Data comparison.} As data we use stellar line-of-sight rotation velocities $v_\text{rot} \equiv \langle v_\text{los} \rangle$ [TO DO: consistent, los or rot] and velocity dispersions $\sigma$ as described in \S\ref{sec:data}. The JAM models give us a prediction for the second line-of-sight velocity moment $v_\text{los}$. The root mean square (rms) line-of-sight velocity $v_\text{rms}$ allows a data-model comparison by relating theses velocities according to 
\begin{equation*}
v_\text{rms}^2 = \langle v_\text{los}^2 \rangle = v_\text{rot}^2 + \sigma^2.
\end{equation*}
The model in Eq. [TO DO] predicts the intrinsic $\langle v_\text{los}^2 \rangle$ at a given position on the sky, which have then to be modified to model the mode of observation, to be comparable to the measurements. The measured $v_\text{rms}$ is a light-weighted mean for a pixel along the long-slit of the spectrograph, with height $L_y = 1$ arcsec \citep{SWELLSV} and a certain given extent in along the major axis, $L_x$, i.e. for a rectangular aperture
\begin{equation*}
\text{AP}(x,y) = \left\{ \begin{array}{ll} 1 & \text{for } -\frac{L_x}{2} \leq x \leq + \frac{L_x}{2} \text{ and } - \frac{L_y}{2} \leq y \leq + \frac{L_y}{2}  \\ 0 & \text{otherwise} \end{array} \right. .
\end{equation*}
The light arriving at the spectrograph itself was subject to seeing, i.e. a Gaussian 
\begin{equation*}
\text{PSF}(x,y)=\mathscr{N}(0,FWHM/2\sqrt{2\ln2})
\end{equation*}
with FWHM=1.1 arcsec  \citep{SWELLSV}. The model predictions have therefore to be convolved with the convolution kernel
\begin{eqnarray*}
K(x,y) &=& (\text{PSF} \ast \text{AP})(x,y) \\
&=& \frac{1}{4} \prod_{u \in \{x,y\}} \left[ \text{erf}\left( \frac{L_u/2 - u}{\sqrt{2}\sigma_\text{seeing}}\right) + \text{erf} \left( \frac{L_u/2 + u}{\sqrt{2} \sigma_\text{seeing}} \right) \right]
\end{eqnarray*}
and weighted by the surface brightness $I(x,y)$ [TO DO: primed x and y or not????]
\begin{eqnarray*}
I_\text{obs} &=& I \ast K\\
\langle v_\text{los}^2 \rangle |_\text{obs} &=& \frac{(I \langle v_\text{los}^2\rangle) \ast K}{I_\text{obs}}.
\end{eqnarray*}
If provided with with the convolution kernel, the JAM code by \citet{Cap08} [TO DO: reference code] performs the convolution numerically. We set $L_x = 0.21$ arcsec as the width of the model pixel, and get a prediction for the actual measurements in bins of width 0.63, 1.26 and 1.89 arcsec \citep{SWELLSV} as light-weighted mean from each 3, 6 and 9 model pixels.

\paragraph{Rotation curve.} The intrinsic rotation curve is the first velocity moment $\langle v_\phi\rangle = \sqrt{\langle v_\phi^2 \rangle - \sigma_\phi^2}$. The observed rotation curve is the projection of the light-weighted contributions to $\langle v_\phi\rangle$ along the line-of-sight(\cite{Cap08}),
\begin{equation*}
I \langle v_\text{los}\rangle = \int_{-\infty}^{+\infty} \nu \langle v_\phi\rangle \cos \phi \sin i \diff z'.
\end{equation*}
The first velocity moments cannot be uniquely determined from the Jeans equations, which give only a prediction for the second velocity moments. Further assumptions are needed to separate the second velocity moments into ordered and random motion. \citet{Cap08} assumes that in a steady state there is no streaming velocity in $R$ direction, i.e. $\langle v_R \rangle = 0$ and therefore $\sigma_R^2 = \langle v_R^2 \rangle$. Then \citet{Cap08} relates the dispersions in $R$ and $\phi$ direction such that
$$\langle v_\phi\rangle = \sqrt{\langle v_\phi^2 \rangle - \sigma_\phi^2} \equiv \kappa \sqrt{\langle v_\phi^2 \rangle - \langle v_R^2 \rangle},$$
and the $\kappa$ parameter quantifies the rotation: $\kappa = 0$ means no rotation at all and $|\kappa| = 1$ describes a velocity dispersion ellipsoid that is a circle in the $R$-$\phi$ plane \citep{Cap08}. The sign of $\kappa$ determines the rotation direction. We can assign a constant $\kappa_k$ to every Gaussian in the MGE formalism and
\begin{equation*}
\nu \langle v_\phi\rangle = \left[\nu \sum_{k} \kappa_k^2 \left( [\nu\langle v_\phi^2 \rangle]_k - [\nu\langle v_R^2 \rangle]_k\right) \right]^{1/2}
\end{equation*} 
is then the light-weighted circular velocity curve, given the second velocity moments found from the Jeans equations.
\\To model the counter-rotating core of J1331 with one free parameter for, we employ the condition that the overall $\kappa(R)$ profile should smoothly and relatively steeply transition from $\kappa(R) = -\kappa' < 0$ at small $R$ [TO DO: Check, ich glaube das hier ist das intrinsische R in Zylinder-Coordinaten] trough $\kappa(R_0) \equiv 0$ and increase to $\kappa(R) = \kappa' > 0$ at large $R$. Our imposed profile is
\begin{equation}
\kappa(R) = \kappa' \frac{R^2 - R_0^2}{R^2 + R_0^2}. \label{eq:kappa_profile}
\end{equation}
We find $\kappa'$ by matching the model $\langle v_\text{los} \rangle$ with the symmetrized $v_\text{rot}$ data, where for a given $\kappa'$ the $\kappa_k$ are found from fitting the MGE generated profile $\kappa(R) = \sum_k \kappa_k \nu_k(r)/\sum_k \nu_k(r)$ to Eq. (\ref{eq:kappa_profile}). The observed zero point is at $R'_0\approx 2$ arcsec. In the deprojected galactic plane the radius of zero rotation would be at a $R_0 \gtrsim 2$ arcsec, and we choose it to be at 2.2 arcsec.

### 2015-08-20 docs/results_JAMwithNFW_v1.01.tex
Wrote section on rotation curve recovery.

\paragraph{Result.} Figure \ref{fig:modelB4_triangle} shows the posterior probability distribution of the fit sampled with an MCMC. Overplotted are also the priors used to constrain the NFW halo. The best fit parameters are summarized in Table \ref{tab:modelB4_bestfit}. We find that the best fit NFW halo strives to be more massive and with a higher concentration than  proposed by the prior. At the same time the model prefers again very negative velocity anisotropies; the sample with the highest probability is pegged at the lower limit of the prior. Both, the concentrated halo and the low anisotropy are needed to reproduce the central dip of the $v_\text{rms}$ curve.

@@

\subsubsection{Predicting the Rotation Curve at larger Radii}

[TO DO]

\begin{figure*}
\centering
\includegraphics[width=0.7\linewidth]{fig/B4_rms_rot_curves_best_model.ps}
\caption{??? Preliminary crappy caption: Model: NFW halo and constant velocity anisotropy, using the the mean / peak values of the MCMC result in the above pdf for the model parameters. Fitting one more free parameter to the rotation curve in the inner regions, predicting the rotation curve and dispersion at larger radii. Green dots are gas kinematics. ??? [TO DO: nice caption]}
\label{fig:???}
\end{figure*}

>>>>

\paragraph{Result.} Figure \ref{fig:modelB4_triangle} shows the posterior probability distribution of the fit sampled with an MCMC. Overplotted are also the priors used to constrain the NFW halo. The best fit parameters are summarized in Table \ref{tab:modelB4_bestfit}. We find that the best fit NFW halo strives to be more massive and with a higher concentration than  proposed by the prior. At the same time the model prefers again very negative velocity anisotropies; the sample with the highest probability is pegged at the lower limit of the prior. Both, the concentrated halo and the low anisotropy are needed to reproduce the central dip of the $v_\text{rms}$ curve. [TO DO: Rewrite. Not very good.]
The first panel in Figure \ref{fig:modelB4_vrot} shows the mean best fit values from Table \ref{tab:modelB4_bestfit}.

@@

\paragraph{Rotation curve.} Following the procedure in \S\ref{sec:model_JAM} we find the rotation curve from the best fit mean model in Table \ref{tab:modelB4_bestfit} by fitting the rotation parameter $\kappa'$ to the symmetrized $v_\text{rot}$ data within $R' = 3.5$ arcsec. The best fit with $\kappa' = 0.76$ is shown in the second panel of Figure \ref{fig:modelB4_vrot}. The third panel gives the dispersion that was calculated simply by $\sigma = \sqrt{v_\text{rms}^2 - v_\text{rot}^2}$. Our assumptions for $\kappa(R)$ nicely reproduce a $v_\text{rot}$ model with counter-rotating core. Although we fitted $v_\text{rot}$ only to the inner regions, the extrapolation to large radii fits the data also very well. While the dispersion $\sigma$ in the center fits by construction quite good, there is a huge discrepancy at large radii. The predicted dispersion is much larger than the data; we would expect the disk rotationally supported and therefore have a low velocity dispersion; especially dispersions as high as $\sim 200 km/s$ are more likely to be observed in the pressure supported bulges of galaxies. There \emph{might} be something unexpeted with the $\sigma$ measurements around $\sim 5$ arcsec, but at large radii the the best fit model NFW halo is simply too massive.

\begin{figure*}
\centering
\includegraphics[width=0.7\linewidth]{fig/B4_rms_rot_curves_best_model.ps}
\caption{Generating the rotation curve from the JAM model with NFW halo and constant velocity anisotropy with the mean parameters in Table \ref{tab:modelB4_bestfit} and with the best fit rotation parameter $\kappa' = 0.76$ in Eq. (\ref{eq:kappa_profile}) (red solid lines). The second velocity moment in the first panel and the first velocity moment in the second panel with the additional fit parameter $\kappa'$ were fitted to the symmetrized data (solid blue points). The velocity dispersion is simply $\sigma = \sqrt{v_\text{rms}^2 - v_\text{rot}^2}$. At larger radii we compare the unsymmetrized data (open blue dots), the gas kinematics from \citet{SWELLSV} (green diamonds) and a JAM model using the same model parameters but the light distribution MGE generated from the Ellipse model in Figure [TO DO: REF FIGURE] as a prediction for the outer regions of J1331 (red dashed lines). The central regions are very well reproduced and we can also greatly predict the rotation curve at larger radii. Only at larger radii the $v_\text{rms}$ and $v_\text{rot}$ overestimate the measurements, probably due to a too massive NFW halo.}
\label{fig:modelB4_vrot}
\end{figure*}
